---
title: "Course Project - ML"
author: "Rafael Vanhoz"
date: "12/09/2020"
output:
  pdf_document: default
  html_document: default
---

Practice Machine learning project
Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here

library

```{r}
library(caret)
library(randomForest)
library(rpart)
library(rattle)
```

DATA

```{r}
downloadcsv <- function(url, nastrings) {
    temp <- tempfile()
    download.file(url, temp, method = "curl")
    data <- read.csv(temp, na.strings = nastrings)
    unlink(temp)
    return(data)
}

trainurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
train.data.raw <- downloadcsv(trainurl, c("", "NA", "#DIV/0!"))

testurl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
test.data.raw <- downloadcsv(testurl, c("", "NA", "#DIV/0!"))
```

The training data has 19622 observations and 160 features, and the distribution of the five measured stances A,B,C,D,E is:

```{r}
dim(train.data.raw)
table(train.data.raw$classe)
```


```{r}
# Drop the first 7 columns as they're unnecessary for predicting.
train.data.clean1 <- train.data.raw[,8:length(colnames(train.data.raw))]
test.data.clean1 <- test.data.raw[,8:length(colnames(test.data.raw))]
# Drop colums with NAs
train.data.clean1 <- train.data.clean1[, colSums(is.na(train.data.clean1)) == 0] 
test.data.clean1 <- test.data.clean1[, colSums(is.na(test.data.clean1)) == 0] 
# Check for near zero variance predictors and drop them if necessary
nzv <- nearZeroVar(train.data.clean1,saveMetrics=TRUE)
zero.var.ind <- sum(nzv$nzv)
if ((zero.var.ind>0)) {
        train.data.clean1 <- train.data.clean1[,nzv$nzv==FALSE]
}
```


The training data is divided into two sets.  This first is a training set with 70% of the data which is used to train the model.  The second is a validation 
set used to assess model performance.  

```{r}
in.training <- createDataPartition(train.data.clean1$classe, p=0.70, list=F)
train.data.final <- train.data.clean1[in.training, ]
validate.data.final <- train.data.clean1[-in.training, ]
```

MODEL DEVELOPMENT
### Train the model  
The training data-set is used to fit a Random Forest model because it 
automatically selects important variables and is robust to correlated 
covariates & outliers in general. 5-fold cross validation is used when 
applying the algorithm. A Random Forest algorithm is a way of averaging 
multiple deep decision trees, trained on different parts of the same data-set,
with the goal of reducing the variance. This typically produces better 
performance at the expense of bias and interpret-ability. The Cross-validation 
technique assesses how the results of a statistical analysis will generalize 
to an independent data set. In 5-fold cross-validation, the original sample 
is randomly partitioned into 5 equal sized sub-samples. a single sample 
is retained for validation and the other sub-samples are used as training 
data. The process is repeated 5 times and the results from the folds are 
averaged.

```{r cache=TRUE}
control.parms <- trainControl(method="cv", 5)
rf.model <- train(classe ~ ., data=train.data.final, method="rf",
                 trControl=control.parms, ntree=251)
rf.model
```

ESTIMATE PERFORMANCE 
The model fit using the training data is tested against the validation data.
Predicted values for the validation data are then compared to the actual 
values. This allows forecasting the accuracy and overall out-of-sample error,
which indicate how well the model will perform with other data.  


```{r}
rf.predict <- predict(rf.model, validate.data.final)
confusionMatrix(as.factor(validate.data.final$classe), rf.predict)
accuracy <- postResample(rf.predict, as.factor(validate.data.final$classe))
acc.out <- accuracy[1]
overall.ose <- 
        1 - as.numeric(confusionMatrix(as.factor(validate.data.final$classe), rf.predict)
                       $overall[1])
```


RESULTS
The accuracy of this model is **`r acc.out`** and the Overall Out-of-Sample 
error is **`r overall.ose`**.

RUN THE MODEL
The model is applied to the test data to produce the results.

```{r}
results <- predict(rf.model, 
                   test.data.clean1[, -length(names(test.data.clean1))])
results
```

DECISION TREE VISUALIZATION

```{r warning=FALSE}
treeModel <- rpart(classe ~ ., data=train.data.final, method="class")
fancyRpartPlot(treeModel)
```
